### 2010-12-20 Documentation/atropos/atropos.tex
ENH: Update to introduction, theory and experiment sections.  Adding Hammers figure.  BWeb results next.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@707 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\doublespacing

@@

\begin{abstract} 
%Neuroanatomical coordinate systems are essential for the
%interpretation of structural and functional imaging studies.
%However, manual delineation of the sulco-gyral complex is time
%consuming and prone to variability given cortical complexity.  This
%work describes an open source, image-based algorithmic approach to
%cortical parcellation which uses training data to propagate
%structural labelings to individual images.  Along with other
%neuroinformatics tools publicly distributed in our ANTs (Advanced Normalization Tools)%
We introduce {\em Atropos}, an ITK-based multivariate $n$-class
Expectation Maximization (EM) segmentation algorithm based on either parametric or non-parametric finite mixture modelings (FMM) of the
class intensities.  Atropos, available in ANTs
\footnote{
http://www.picsl.upenn.edu/ANTs
}, is capable of incorporating spatial prior probability maps,
(sparse) prior label maps and/or Markov Random Field (MRF) modeling
assumptions.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 68 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) Markov Random
Field Segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based Segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 68 class EM segmentation
problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
illustrative examples that excercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

@@

\section{Introduction} As medical image acquisition technology has advanced,
concomitant investment has been made towards adapting classification
techniques for neuroanatomy.  Some of the earliest work appropriated NASA
satellite image processing software for statistical classification of
head tissues in 2-D MR images \cite{Vannier1985}.  A proliferation of
techniques ensued with increasing sophistication in both core
methodology and degree of refinement for specific problems.  The
chronology of progress in segmentation may be tracked through both
technical reviews
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
and evaluation studies
\citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

@@

EM algorithms are sensitive to local optima. Consequently, a good initialization
strategy is critical.  Common low-level initialization strategies include uniform
probability assignment \citep{Wells1996}, Otsu thresholding
\citep{Zhang2001}, and K-means clustering \citep{Pappas1992}.  More
sophisticated low-level initialization schemes include that of
\cite{Greenspan2006} in which a dense spatial distribution of
Gaussians is used to capture the complex neuroanatomical layout with
subsequent processing used to conjoin subsets of Gaussians belonging
to the same tissue classes.  Recently, anatomically specific
strategies for incorporating prior knowledge into the segmentation
solution have included the use of spatial prior probability maps of
the various structures of interest
\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
probability maps can also be used to provide an initial segmentation.
Related technological developments encompasses those driven by partial
volume modeling for increased accuracy in brain segmentation
\citep{Ruan2000,Ballester2002,Leemput2003}.

A general trend towards more integrative neuroanatomical image processing led to the work described in
\cite{Ashburner2005} which is publicly available within SPM5---a
large-scale Matlab module in which registration, segmentation, and
bias field correction can be simultaneously modeled within a single
optimization scheme. The roots of this very popular software package
stem back to early work by Karl Friston in which the basis for
statistical parametric mapping was developed \citep{Friston1990} and
subsequently provided to the research community. A similar holistic
approach to brain processing was provided in \cite{Pohl2006} in which
segmentation and registration parameters were optimized simultaneously
while casting the inhomogeneity model parameters of \cite{Wells1996}
as nuisance variables.  Continued work involved recursive parcellation
of the brain volume by considering sub-structures in a hierarchical
manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
\citep{Pieper2006}---an open source package with developmental
contributions from multiple agencies including both private and
academic institutions. Researchers in aging often focus on
accurately segmenting the T1 MRI of elderly controls and subjects
suffering from neurodegeneration, for instance, via the SIENA algorithm
\citep{Smith2007}.  Two recent comparative evaluation studies \citep{,Bresser2011} have shown that methods such as FAST,
SPM Unified Segmentation, SIENA and others show different performance
characteristics under different evaluation criteria.  That is, no
single method performs best under every measurement, highlighting
the need for segmentation tools that are tunable for different
problems and research goals.

@@

Our open source segmentation tool, which we have dubbed {\em
Atropos}, 
\footnote{ Atropos is one of the three Fates from Greek
mythology characterized by her dreaded shears used to decide the
destiny of each mortal.  Also, consistent with the entomological motif
of our ANTs toolkit, {\it Acherontia atropos} is a species of large
moth known for the skull-like pattern visible on its thorax.  }
efficiently and flexibly implements a $n$-tissue paradigm for
voxel-based image segmentation.  Atropos allows users to harness its
generalized EM algorithm for standard tissue classification of
the brain into gray matter, white matter and cerebrospinal fluid.
Atropos equally allows incarnations that use EM to simultaneously
maximize the posterior probabilities of many classes, for instance,
when parcellating the brain into hemispheres, cortical regions and
deep brain structures such as amygdala, hippocampus and
thalamus.  Atropos contains features of its predecessors for
performing $n$-tissue segmentation including imposition of prior
information in the form of MRFs and template-based spatial prior
probability maps as well as weighted combinations of these terms.  We
extend the use of spatial priors, in this work, to allow sparse priors
that may then be extended by a geodesic distance-based label propagation
approach.  In short, Atropos seeks to provide a toolbox of
segmentation approaches that may be modified, tuned and refined by the user
for different problems.

@@

The major contribution of the Atropos algorithm is a flexible,
well-designed, memory efficient implementation of core expectation maximization machinery
that is useful across many segmentation problems.  Atropos options
include different maximization criterion, an interactive mode, tunable
priors and multivariate density estimators using both
non-parametric and parametric functions.  To highlight the
value of this open-source, free contribution, we performed a search of
software attributes on NITRC and found that as of November 2010 no
stand-alone EM methods are currently listed.
Atropos's memory efficiency enables EM-segmentation of many classes at
once, a capability that is due to a sparse internal representation of
prior probability images that minimizes the need for random access
memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
neuroimages into 68 neuronatomical regions to illustrate the practical
value of the low-memory implementation within this paper.  Atropos
also allows multivariate image features to drive the data term, relevant when more than one view of anatomy aids
segmentation, as in neonatal brain tissue classification
\cite{Weisenfeld2009}.  Although Atropos may be applied to
multivariate data from arbitrary modalities, we limit our evaluation
to tissue-classification in T1 neuroimaging in part due to the
abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

@@

Bayes' theorem provides a powerful mechanism for making inductive
inferences assuming the availability of quantities defining the
relevant conditional probabilities, specifically the likelihood and
prior probability terms.  Bayesian paradigms for brain image
segmentation utilize a user-selected observation model defining the
likelihood term and one or more prior probability terms.  The produce
of likelihood(s) and prior(s) is proportional to the posterior
probability.  The likelihood term has been previously defined both 
parametrically (e.g. a Gaussian model) and non-parametrically
(e.g. Parzen windowing of the sample histogram).  The prior term, as
given in the literature, has often been formed either as
MRF-based or template-based.  An image segmentation solution in this
context is an assignment of one label to each voxel%
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  The next sections
introduce notation and provide a formal description of three essential
components in Bayesian segmentation, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized
    MRF and template-based prior terms, and 
%, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}
These components are common across most EM segmentation approaches.

@@

\subsection{Likelihood or Observation Models}
To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normal distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
the mean and variance of the $k^{th}$ model.  In the case of
multi-modal data, where $y_i$ is a vector quantity, a mulitvariate
Gaussian is defined by the mean vector, $\boldsymbol{\mu}_k$, and
covariance matrix, $\Sigma_k$.  Multivariate probabilities are derived from the
Mahalanobis distance.  An initial estimate for these model
parameters are tabulated during the initialization process,
e.g. K-means or Otsu thresholding.

@@

Atropos enables two choices for the likelihood formulation.  One is
based on the classical finite mixture model (FMM) and the second is based on
the spatially varying mixture model (SVMM) defined in \citep{Sanjay-Gopal1998}.
In order to calculate the probability associated with the entire set of observations, $\mathbf{y}$, we assume independency between voxels which is done primarily for facilitating quantification as opposed to modeling realistic assumptions.  Spatial interdependency between voxels is modeled by the prior probabilities discussed in the next section.  Marginalizing over the set of possible labels, $\mathcal{L}$, leads to the following probabilistic formulation
\begin{align}\label{eq:likelihood}
  p(\mathbf{y}|\mathbf{x}) = \prod_{i=1}^N \left(      
                                                        \sum_{k=1}^K \gamma_k p_k(y_i|l_k)
                                                        \right)
\end{align}
where $\gamma_k$ is the mixing parameter \citep{Ashburner2005}.

\subsection{Prior Probability Models}
By modeling $\mathcal{F}$ via the set of observation models $\Phi$,
this so called finite-mixture model could be used to produce a
labeling or segmentation \citep[e.g.][]{Wells1996}.  However, as
pointed out by \cite{Zhang2001}, the exclusive use of the observation
models in modeling the intensity profile produces a less than optimal
solution because spatial contextual considerations are ignored.  This
has been remedied by the introduction of a host of prior probability
models including those characterized by use of MRF theory and
template-based information.  For example, in the works of
\cite{Leemput1999a} and \cite{Weisenfeld2009}, the original global
prior term given in \cite{Wells1996} is replaced by the product of the
template-based and the MRF-based prior terms.  In addition to their
descriptions below, we discuss our introduction of a third possible
prior/objective combination in the form of a (sparse) prior labeling which fixes specific points
of the segmentation and uses EM to propagate this information
elsewhere in the image.  
% extension of template-based Euclidean and geodesic distance label propagation.

@@

We employ the strategy given in \cite{Ashburner2005} in which the
stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$,
FIXME DEFINE THE GAMMA --- SHOULD USE SVMM FROM HEBERT ?
describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}
\end{align}
where $t_{ik}$ is the prior probability value due to the
template-based prior information at site $i$.  The user may also
choose mixing proportions equal to $\frac{1}{K}$ via the command line
interface to the posterior formulation.

% {\textcolor{red}{FIXME --- need explanation of heuristic ...  Also, we

%  allow a weighted combination defined by the ComputeLocalPosterior

%  function in Atropos --- how do we justify / explain this?   clearly,

% it's a balance of the global solution (no local prior) and the local

% one with spatial prior ... }}

\subsubsection{Supervised Initialization to the EM Algorithm}
PriorLabelImage

@@

\subsection{Optimization}
Atropos uses expectation maximization to find a locally optimal
solution for the user-selected version of the Bayesian segmentation
problem (cf. Eqn. (\ref{eq:bayes})).
After initial estimation of the
likelihood model parameters, EM iterates between calculation of the
missing $\hat{\mathbf{x}}$ and subsequent re-estimation of the model
parameters by maximizing the expectation of the complete data
log-likelihood (cf. Eqn. (\ref{eq:likelihood})).  
%\begin{align}
%\log( p(\mathbf{y}) ) = \sum_{i=1}^N \log \left( \sum_{k=1}^K \gamma_k  p(\mathbf{y}|l_k)  \right)
%\end{align}
The expectation maximization procedure is derived in various
publications including \cite{Zhang2001} which yields the optimal mean
and variance (or covariance), but sets the mixing parameter $\gamma_k$
as a constant.  The Atropos implementation estimates $\gamma_k$ at
each iteration as in \cite{Ashburner2005}.
\footnote{
Due to the lack of parameters in the non-parametric approach, it is not technically an EM algorithm (as described in \cite{Wells1996}).  However, the same iterative strategy is applicable and is quite robust in practice as evidenced by the number of researchers employing non-parametric strategies (see the Introduction).
}  
When spatial coherence constraints are included as an MRF prior in Atropos, the optimal segmentation solution becomes intractable.%
\footnote{
Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.  Due to the large $K$ associated with $K>>1$ problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
}
Although many optimization techniques exist \citep[see the
introduction in][for a concise summary of the myriad optimization
possibilities]{Marroquin2002}---each with their characteristic
advantages and disadvantages in terms of computational complexity and
accuracy---Atropos uses the well-known {\em Iterated Conditional
  Modes} (ICM)  \citep{Besag1986} which is greedy, computationally
efficient and provides good performance.  The EM algorithm employed in
Atropos may therefore be written
\begin{eqnarray}
X \\ 
Y \\
Z
\end{eqnarray}
The solution will be a local maxima.

%\begin{align}
%P(l_k | y_i, t_i, m_i,d_i) &= \frac{P(y_i|l_k ,t_i,m_i,d_i)   P(l_k|t_i,m_i,d_i)   }{P(y_i|t_i,m_i,d_i)}
%\end{align}

>>>>

% \doublespacing

@@

\begin{abstract} 
%Neuroanatomical coordinate systems are essential for the
%interpretation of structural and functional imaging studies.
%However, manual delineation of the sulco-gyral complex is time
%consuming and prone to variability given cortical complexity.  This
%work describes an open source, image-based algorithmic approach to
%cortical parcellation which uses training data to propagate
%structural labelings to individual images.  Along with other
%neuroinformatics tools publicly distributed in our ANTs (Advanced Normalization Tools)%
We introduce {\em Atropos}, an ITK-based multivariate $n$-class
Expectation Maximization (EM) segmentation algorithm based on either parametric or non-parametric finite mixture modeling (FMM) of the
class intensities.  Atropos, available in ANTs
\footnote{
http://www.picsl.upenn.edu/ANTs
}, is capable of incorporating spatial prior probability maps,
(sparse) prior label maps and/or Markov Random Field (MRF) modeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 68 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) MRF segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 68 class EM segmentation
problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
illustrative examples that excercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

@@

\section{Introduction} As medical image acquisition technology has advanced,
concomitant investment has been made towards adapting classification
techniques for neuroanatomy.  Some of the earliest work appropriated NASA
satellite image processing software for statistical classification of
head tissues in 2-D MR images \citep{Vannier1985}.  A proliferation of
techniques ensued with increasing sophistication in both core
methodology and degree of refinement for specific problems.  The
chronology of progress in segmentation may be tracked through both
technical reviews
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
and evaluation studies
\citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

@@

EM algorithms are sensitive to local optima. Consequently, a good initialization
strategy is critical.  Common low-level initialization strategies include uniform
probability assignment \citep{Wells1996}, Otsu thresholding
\citep{Zhang2001}, and K-means clustering \citep{Pappas1992}.  More
sophisticated low-level initialization schemes include that of
\cite{Greenspan2006} in which a dense spatial distribution of
Gaussians is used to capture the complex neuroanatomical layout with
subsequent processing used to conjoin subsets of Gaussians belonging
to the same tissue classes.  Recently, anatomically specific
strategies for incorporating prior knowledge into the segmentation
solution have included the use of spatial prior probability maps of
the various structures of interest
\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
probability maps can also be used to provide an initial segmentation.
Related technological developments model partial
volume effects for increased accuracy in brain segmentation
\citep{Ruan2000,Ballester2002,Leemput2003}.

A general trend towards more integrative neuroanatomical image processing led to the work described in
\cite{Ashburner2005} which is publicly available within SPM5, a
large-scale Matlab module in which registration, segmentation, and
bias field correction can be simultaneously modeled within a single
optimization scheme. The roots of this very popular software package
stem back to early work by Karl Friston in which the basis for
statistical parametric mapping was developed \citep{Friston1990} and
subsequently provided to the research community. A similar holistic
approach to brain processing was provided in \cite{Pohl2006} in which
segmentation and registration parameters were optimized simultaneously
while casting the inhomogeneity model parameters of \cite{Wells1996}
as nuisance variables.  Continued work involved recursive parcellation
of the brain volume by considering sub-structures in a hierarchical
manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
\citep{Pieper2006}---an open source package with developmental
contributions from multiple agencies including both private and
academic institutions. Researchers in aging often focus on
accurately segmenting the T1 MRI of elderly controls and subjects
suffering from neurodegeneration, for instance, via the SIENA algorithm
\citep{Smith2007}.  A recent evaluation study 
compared kNN segmentation, SPM Unified Segmentation and SIENA and
found different performance
characteristics under different evaluation criteria
\citep{Bresser2011}.  \cite{Klauschen2009} had similar findings when
comparing SPM5, FSL and FreeSurfer.  These studies suggest that no
single method performs best under every measurement and, along with the No
Free Lunch theorem, highlight the need for segmentation tools that are tunable for different
problems and research goals.

@@

Our open source segmentation tool, which we have dubbed {\em
Atropos}, 
\footnote{ Atropos is one of the three Fates from Greek
mythology characterized by her dreaded shears used to decide the
destiny of each mortal.  Also, consistent with the entomological motif
of our ANTs toolkit, {\it Acherontia atropos} is a species of large
moth known for the skull-like pattern visible on its thorax.  }
efficiently and flexibly implements a $n$-tissue paradigm for
voxel-based image segmentation.  Atropos allows users to harness its
generalized EM algorithm for standard tissue classification of
the brain into gray matter, white matter and cerebrospinal fluid.
Atropos equally allows incarnations that use EM to simultaneously
maximize the posterior probabilities of many classes, for instance,
when parcellating the brain into hemispheres, cortical regions and
deep brain structures such as amygdala, hippocampus and
thalamus.  Atropos contains features of its predecessors for
performing $n$-tissue segmentation including imposition of prior
information in the form of MRFs and template-based spatial prior
probability maps as well as weighted combinations of these terms.  We
also borrow an idea from \cite{Boykov2004} and use sparse spatial
priors to provide initialization and boundary conditions for Atropos
EM segmentation in a semi-interactive manner.  In short, Atropos seeks
to provide a toolbox of segmentation approaches that may be modified,
tuned and refined by the user for different problems.

@@

The Atropos algorithm is multi-platform, free, stand-alone and able to
manage a wide range of scalar and vector image datatypes.  To highlight the
value of this open-source contribution, we performed a search of
software attributes on NITRC and found that as of November 2010 no
stand-alone EM methods are currently listed.
Atropos's memory efficiency enables EM-segmentation of many classes at
once, a capability that is due to a sparse internal representation of
prior probability images that minimizes the need for random access
memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
neuroimages into 68 neuronatomical regions to illustrate the practical
value of the low-memory implementation within this paper.  Atropos
also allows multivariate image features to drive the data term, relevant when more than one view of anatomy aids
segmentation, as in neonatal brain tissue classification
\cite{Weisenfeld2009}.  Although Atropos may be applied to
multivariate data from arbitrary modalities, we limit our evaluation
to tissue-classification in T1 neuroimaging in part due to the
abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

@@

Bayes' theorem provides a powerful mechanism for making inductive
inferences assuming the availability of quantities defining the
relevant conditional probabilities, specifically the likelihood and
prior probability terms.  Bayesian paradigms for brain image
segmentation employ a user-selected observation model defining the
likelihood term and one or more prior probability terms.  The product
of likelihood(s) and prior(s) is proportional to the posterior
probability.  The likelihood term has been previously defined both 
parametrically (e.g. a Gaussian model) and non-parametrically
(e.g. Parzen windowing of the sample histogram).  The prior term, as
given in the literature, has often been formed either as
MRF-based or template-based.  An image segmentation solution in this
context is an assignment of one label to each voxel
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  The next sections
introduce notation and provide a formal description of three essential
components in Bayesian segmentation, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized
    MRF and template-based prior terms, and 
%, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}
These components are common across most EM segmentation approaches.

@@

\subsection{Likelihood or Observation Models}
To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normal distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
the mean and variance of the $k^{th}$ model.  When $y_i$ is a vector
quantity, we replace the Euclidean distance by by Mahalanobis distance
and define multivariate Gaussian parameters via a mean vector,
$\boldsymbol{\mu}_k$, and covariance matrix, ${\bf \Sigma_k}$.

@@

Atropos segmentation likelihood estimates are based on the classical finite mixture model (FMM).
FMM assumes independency between voxels to calculate the probability
associated with the entire set of observations, $\mathbf{y}$.  Spatial
interdependency between voxels is modeled by the prior probabilities
discussed in the next section.  Marginalizing over the set of possible labels, $\mathcal{L}$, leads to the following probabilistic formulation
\begin{align}\label{eq:likelihood}
  p(\mathbf{y}|\mathbf{x}) = \prod_{i=1}^N \left(      
                                                        \sum_{k=1}^K \gamma_k p_k(y_i|l_k)
                                                        \right)
\end{align}
where $\gamma_k$ is the mixing parameter \citep{Ashburner2005}.

\subsection{Prior Probability Models}
By modeling $\mathcal{F}$ via the set of observation models $\Phi$,
this so called finite-mixture model could be used to produce a
labeling or segmentation \citep[e.g.][]{Wells1996}.  However, as
pointed out by \cite{Zhang2001}, exclusive use of the intensity profile produces a less than optimal
solution because spatial contextual considerations are ignored.  This
has been remedied by the introduction of a host of prior probability
models including those characterized by use of MRF theory and
template-based information.  For example, in the works of
\cite{Leemput1999a} and \cite{Weisenfeld2009}, the original global
prior term given in \cite{Wells1996} is replaced by the product of the
template-based and the MRF-based prior terms.  In addition to their
descriptions below, we discuss a third possible
prior/objective combination in the form of a (sparse) prior labeling which fixes specific points
of the segmentation and uses EM to propagate this information
elsewhere in the image.  
% extension of template-based Euclidean and geodesic distance label propagation.

@@

We employ the strategy given in \cite{Ashburner2005} in which the
stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$
(cf. Eqn. (\ref{eq:likelihood})), 
describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions,
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}.
\end{align}
The $t_{ik}$ is the prior probability value at site $i$ which was
mapped, typically by image registration, to the local image
from a template data set.  The user may also
choose mixing proportions equal to
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{t_{ik}}{\sum_{j=1}^K t_{ij}}.
\end{align}
via the command line interface to the posterior formulation.
{\textcolor{red}{FIXME --- need explanation of heuristic ... how do we
  estimate the $\gamma$ without the $t_{ij}$ present?}

\subsubsection{Supervised Semi-Interactive Segmentation}
Brain segmentation methods have relied on user interaction for many
years \citep{Lim1989,Julin1997,Freeborough1997a,Yushkevich2006}.
Atropos is capable of benefitting from user knowledge via an
initialization and optimization strategy that depends upon a spatially
varying prior label image passed as input.  Rapid, sparse
labeling---with visualization provided by ITK-SNAP
(www.itksnap.org)---enables an interaction and execution processing
loop that can be critical to solving segmentation problems with
challenging clinical data in which automated approaches fail.  This
part of Atropos design is inspired by the interactive graph cut methods
pioneered by \cite{Boykov2001} and which has spawned many follow-up
applications.  The Atropos prior label
image pre-specificies the segmentation results at a subset of the
spatial domain by fixing the priors and likelihood (and, thus, the
posterior) at a subset of $\mathcal{I}$ to be $1$ for the known label
and $0$ for each other label at the same site.  The user-input
therefore not only initializes the algorithm, but also gives boundary conditions
that influence the EM solution outside of the known sites.  While the
graph-based min-cut max-flow solution is globally optimal for two
labels, only locally optimal approachs are available for 3 or more
classes.  Thus, in most practical applications, the EM approach is a
reasonable and efficient alterative to Boykov's solution.
Furthermore, as with similar algorithms, one may automate the
initialization process.

@@

\subsection{Optimization}
Atropos uses expectation maximization to find a locally optimal
solution for the user-selected version of the Bayesian segmentation
problem (cf. Eqn. (\ref{eq:bayes})).
After initial estimation of the
likelihood model parameters, EM iterates between calculation of the
missing optimal labels $\hat{\mathbf{x}}$ and subsequent re-estimation of the model
parameters by maximizing the expectation of the complete data
log-likelihood (cf. Eqn. (\ref{eq:likelihood})).  
%\begin{align}
%\log( p(\mathbf{y}) ) = \sum_{i=1}^N \log \left( \sum_{k=1}^K \gamma_k  p(\mathbf{y}|l_k)  \right)
%\end{align}
The expectation maximization procedure is derived in various
publications including \cite{Zhang2001} which yields the optimal mean
and variance (or covariance), but sets the mixing parameter $\gamma_k$
as a constant.  The Atropos implementation estimates $\gamma_k$ at
each iteration as in \cite{Ashburner2005}.
\footnote{
Due to the lack of parameters in the non-parametric approach, it is not technically an EM algorithm (as described in \cite{Wells1996}).  However, the same iterative strategy is applicable and is quite robust in practice as evidenced by the number of researchers employing non-parametric strategies (see the Introduction).
}  
When spatial coherence constraints are included as an MRF prior in Atropos, the optimal segmentation solution becomes intractable.%
\footnote{
Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.  Due to the large $K$ associated with $K>>1$ problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
}
Although many optimization techniques exist \citep[see the
introduction in][for a concise summary of the myriad optimization
possibilities]{Marroquin2002}---each with their characteristic
advantages and disadvantages in terms of computational complexity and
accuracy---Atropos uses the well-known {\em Iterated Conditional
  Modes} (ICM)  \citep{Besag1986} which is greedy, computationally
efficient and provides good performance.  The EM algorithm employed in
Atropos may therefore be written as a series of steps:
\begin{description}
\item[Initialization:] In all cases, the user defines the number of
classes to segment.  The simplest initialization is by the classic
K-means algorithm with only the number of classes as a prior.
Otherwise, the user must provide prior information for each class in
the form of either a single $n$-ary prior label image or a series of
prior probability images, one for each class.
\item[E-Step: ICM Label Update:] Given the initialization and fixed
model parameters, Atropos will evaluate the posterior probability for
each possible label and assign, to each voxel, the label that
maximizes the posterior.  {\bf \textcolor{red}{ FIXME more specific?}}.
\item[M-Step: Parameter Update:] We describe the parametric update.  Given fixed labels, we update the
  parameters by the FMM-defined likelihood.  We compute, for each of
  $K$ labels, the mean,
$$ \mu^{new}_k \leftarrow \frac{ \sum_{i=1}^N  y_i \gamma_k p_k(y_i|l_k)}{ \sum_{i=1}^N \gamma_k p_k(y_i|l_k) },
$$
variance {\bf \textcolor{red}{ FIXME should this be the full
    covariance matrix?}},
$$ \sigma^{new}_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k ) \gamma_k p_k(y_i|l_k)}{ \sum_{i=1}^N  \gamma_k p_k(y_i|l_k) },
$$
and mixture parameters, 
$$ \gamma^{new}_k \leftarrow  \textcolor{red}{ FIXME .}
$$
This type of update strategy is known as {\em soft} EM.  Hard EM, in
contrast, only uses sites containing label $l_k$ to update model $k$ parameters.
A similar strategy is used in the multivariate and non-parametric cases.
\end{description}
The EM algorithm will iterate toward a local maximum.  We track
convergence of the algorithm by summing up the maximum posterior
probability at each site over the segmentation domain.  
%\begin{align}
%P(l_k | y_i, t_i, m_i,d_i) &= \frac{P(y_i|l_k ,t_i,m_i,d_i)   P(l_k|t_i,m_i,d_i)   }{P(y_i|t_i,m_i,d_i)}
%\end{align}
%
%
%An image, $I$, maps a domain, $\Omega$ into the
%positive real numbers, such that $I \colon \Omega \rightarrow
%\mathbb{R}^+$.  The goal of segmentation, in general, is to define the
%spatial distribution of a finite set of labels over this domain.  We
%denote the segmentation itself as $\eta \colon \Omega \rightarrow L$
%where $L = \{ L_1 = 1 , \cdots , L_N=N \}$, a set of integer indexed
%segmentation labels.  Note that $\eta$ may be formed as $\eta =
%\sum_{i=1}^{i=N} L_i \eta_i $ where $\eta_i$ is the binary
%segmentation for label $L_i$.  A prior estimate for the label image
%$\eta_i$ is here denoted $\eta^s_i$ with a complete set of priors
%denoted $\eta^s$.  The boundary of the binary segmentation -- where a
%$0/1$ transition edge exists -- is denoted $\partial \eta^s_i$, for
%the prior, and $\partial \eta_i$ for the label image.
%
%\subsection{Apocrita Theory} A general maximum a posteriori criterion
%for segmentation seeks,
%\begin{equation} {\hat \eta} = \argmax _{\eta} \Pr( \eta | I )(\x) =
%\Pr( I | \eta )(\x) \Pr(\eta)(\x),
%\end{equation} 
%where $I$ is the input image, ${\eta}$ represents the
%label set configuration taken from the set $L$, $\Pr$ is the
%probability, $\x$ is the spatial index and the optimal solution is
%$\hat \eta$.  The input image $I$, here, is an unlabeled T1 MRI
%indexed by the value $\x \in \Omega$ where $\Omega$ is the image's
%spatial domain.  This probability is composed of the likelihood (first
%term) and the prior (second term).  Atropos ~ uses a spatially varying
%likelihood term and a two-component prior term that takes into account
%both spatial distribution and label smoothness, the latter via a
%standard MRF prior.
%
%The likelihood term for a single label value $L_i \in L$ is,
%\begin{eqnarray} \Pr( I | \eta_i )(\x)=\frac{1}{Z_i} \exp(- \| I(\x) -
%\mu_i(\x) \|^2 / \sigma_i^2 ),
